
\subsection{Starting the default R GUI}

Starting R is simple. If you're using Windows simply navigate to the R
subfolder from the Start Menu. On a Unix/Linux system invoke the program by
typing |R|. On \OSX start R by clicking the R icon from the Dock or in the
Applications folder. The \OSX and Windows version of R provide a simple GUI
interface that simplifies certain tasks. When you start up the R GUI you'll be
presented with a single window, the R console. The rest of this document will
assume you're using R under Windows or \OSX.

\subsection{R Studio}

\href{http://rstudio.org/}{RStudio} is a new IDE for R being developed under
an open source model (see the
\href{https://github.com/rstudio/rstudio}{RStudio GitHub Site}). It provides a
nicely designed graphical interface that is consistent across platforms. It
can even run as a server, allowing you to access R via a web interface. Check
out the \href{http://rstudio.org/docs/}{RStudio Docs} for detailed info on
configuring Rstudio. We'll go over some of RStudio's nice features in class.

\subsection{Accessing the Help System on R}

R comes with fairly extensive documentation and a simple help system. You can
access HTML versions of R documentation under the Help menu in the GUI. The
HTML documentation also includes information on any packages you've installed.
Take a few minutes to browse through the R HTML documentation.

The help system can be invoked from the console itself using the
|help| function or the |?| operator.
%
\begin{R}
> help(length)
> ?length 
> ?log
\end{R}
%
What if you don't know the name of the function you want? You can use
the |help.search()| function.
%
\begin{R}
> help.search("log")
\end{R}
%
In this case |help.search("log")| returns all the functions with
the string `log' in them. For more on |help.search| type
|?help.search|. Other useful help related functions include
|apropos()| and |example()|.

\subsection{Navigating Directories in R}

When you start the R environment your `working directory' (i.e.~the
directory on your computer's file system that R currently `sees')
defaults to a specific directory. On Windows this is usually the same
directory that R is installed in, on \OSX it is typically your home
directory. Here are examples showing how you can get information about
your working directory and change your working directory.
%
\begin{R}
> getwd() 
[1] "/Users/pmagwene"
> setwd("/Users") 
> getwd() 
[1] "/Users"
\end{R}
%
Note that on Windows you can change your working directory by using the
|Change dir...| item under the |File| menu, while the corresponding item is found under the |Misc| menu on \OSX.

To get a list of the file in your current working directory use the
|list.files()| function.
%
\begin{R}
> list.files() 
[1] "Shared" "pmagwene"
\end{R}



\subsection{Using R as a Calculator}

The simplest way to use R is as a fancy calculator.
%
\begin{R}
> 3.14 * 2.5^2
[1] 19.625
> pi * 2.5^2 # R knows about some constants such as Pi 
[1] 19.63495
> cos(pi/3)
[1] 0.5
> sin(pi/3)
[1] 0.8660254
> log(10)
[1] 2.302585
> log10(10) # log base 10
[1] 1
> log2(10) # log base 2
[1] 3.321928
> (10 + 2)/(4-5)
[1] -12
> (10 + 2)/4-5 # compare the answer to the above
[1] -2
\end{R}
%
Be aware that certain operators have precedence over others. For example
multiplication and division have higher precedence than addition and
subtraction. Use parentheses to disambiguate potentially confusing
statements.

\begin{R}
> sqrt(pi)
[1] 1.772454
> sqrt(-1)
[1] NaN
Warning message:
NaNs produced in: sqrt(-1) 
> sqrt(-1+0i)
[1] 0+1i
\end{R}
%
What happened when you tried to calculate |sqrt(-1)|?, -1 is
treated as a real number and since square roots are undefined for the
negative reals, R produced a warning message and returned a special
value called |NaN| (Not a Number). Note that square roots of
negative complex numbers are well defined so |sqrt(-1+0i)|
works fine.
%
\begin{R}
> 1/0
[1] Inf
\end{R}
Division by zero produces an object that represents infinite numbers.

\subsection{Comparison Operators}

You've already been introduced to the most commonly used arithmetic
operators. Also useful are the comparison operators:
%
\begin{R}
> 10 < 9  # less than
[1] FALSE
> 10 > 9  # greater than
[1] TRUE
> 10 <= (5 * 2) # less than or equal to
[1] TRUE
> 10 >= pi # greater than or equal to
[1] TRUE
> 10 == 10 # equals 
[1] TRUE
> 10 != 10 # does not equal
[1] FALSE
> 10 == (sqrt(10)^2) # Surprised by the result? See below.
[1] FALSE
> 4 == (sqrt(4)^2) # Even more confused?
[1] TRUE
\end{R}
%
Comparisons return boolean values. Be careful to distinguish between
|==| (tests equality) and |=| (the alternative
assignment operator equivalent to |<-|).

How about the last two statement comparing two values to the square of
their square roots? Mathematically we know that both
$(\sqrt{10})^2 = 10$ and $(\sqrt{4})^2 = 4$ are true statements. Why
does R tell us the first statement is false? What we're running into
here are the limits of computer precision. A computer can't represent
$\sqrt 10$ exactly, whereas $\sqrt 4$ can be exactly represented.
Precision in numerical computing is a complex subject and beyond the
scope of this course. Later in the course we'll discuss some ways of
implementing sanity checks to avoid situations like that illustrated
above.

\subsection{Working with Vectors in R}

\subsubsection{Vector Arithmetic and Comparison}

Remember that in R arithmetic operations work on vectors as well as on
single numbers (in fact single numbers \emph{are} vectors).
%
\begin{R}
> x <- c(2, 4, 6, 8, 10)
> x * 2
[1]  4  8 12 16 20
> x * pi
[1]  6.283185 12.566371 18.849556 25.132741 31.415927
> y <- c(0, 1, 3, 5, 9)
> x + y
[1]  2  5  9 13 19
> x * y
[1]  0  4 18 40 90
> x/y
[1]      Inf 4.000000 2.000000 1.600000 1.111111
> z <- c(1, 4, 7, 11)
> x + z
[1]  3  8 13 19 11
Warning message:
longer object length
        is not a multiple of shorter object length in: x + z 
\end{R}
%
When vectors are not of the same length R `recycles' the elements of the
shorter vector to make the lengths conform. In the example above
|z| was treated as if it was the vector |(1, 4, 7, 11, 1)|.

The comparison operators also work on vectors as shown below.
Comparisons involving vectors return vectors of booleans.
%
\begin{R}
> x > 5
[1] FALSE FALSE  TRUE  TRUE  TRUE
> x != 4
[1]  TRUE FALSE  TRUE  TRUE  TRUE
\end{R}


\subsubsection{Indexing Vectors}

For a vector of length $n$, we can access the elements by the indices $1
\ldots n$. Trying to access an element beyond these limits returns a special
constant called |NA| (Not Available) that indicates missing or non-existent
values.
%
\begin{R}
> length(x)
[1] 5
> x[1]
[1] 2
> x[4]
[1] 8
> x[6]
[1] NA
> x[-1]
[1]  4  6  8 10
> x[c(3,5)]
[1]  6 10
\end{R}
%
Negative indices are used to exclude particular elements. |x[-1]| returns all
elements of |x| except the first. You can get multiple elements of a vector by
indexing by another vector. In the example above |x[c(3,5)]| returns the third
and fifth element of |x|.

\subsubsection{Combining Indexing and Comparison}

A very powerful feature of R is the ability to combine the comparison
operators with indexing. This facilitates data filtering and subsetting.
Some examples:
%
\begin{R}
> x <- c(2, 4, 6, 8, 10)
> x[x > 5]
[1]  6  8 10
> x[x < 4 | x > 6]
[1]  2  8 10
\end{R}
%

In the first example we retrieved all the elements of |x| that are larger than
5 (read as `x where x is greater than 5'). In the second example we retrieved
those elements of |x| that were smaller than four \emph{or} greater than six.
The symbol \lstinline!|! is the `logical or' operator. Other logical operators
include |&| (`logical and' or `intersection') and |!| (negation). Combining
indexing and comparison is a powerful concept and one you'll probably find
useful for analyzing your own data.

\subsubsection{Generating Regular Sequences}

Creating sequences of numbers that are separated by a specified value or
that follow a particular patterns turns out to be a common task in
programming. R has some built-in operators and functions to simplify
this task.
%
\begin{R}
> s <- 1:10
> s
 [1]  1  2  3  4  5  6  7  8  9 10
> s <- 10:1
> s
 [1] 10  9  8  7  6  5  4  3  2  1
> s <- seq(0.5,1.5,by=0.1)
> s
 [1] 0.5 0.6 0.7 0.8 0.9 1.0 1.1 1.2 1.3 1.4 1.5
# 'by' is the 3rd argument so you don't have to specify it
> s <- seq(0.5, 1.5, 0.33) 
> s
[1] 0.50 0.83 1.16 1.49 
\end{R}

|rep()| is another way to generate patterned data.
%
\begin{R}
> rep(c("Male","Female"),3)
[1] "Male"   "Female" "Male"   "Female" "Male"   "Female"
> rep(c(T,T, F),2)
[1]  TRUE  TRUE FALSE  TRUE  TRUE FALSE
\end{R}


\subsection{Some Useful Functions}

You've already seem a number of functions (e.g. |sin()|,
|log|, |length()|, etc). Functions are called by
invoking the function name followed by parentheses containing zero or
more \emph{arguments} to the function. Arguments can include the data
the function operates on as well as settings for function parameter
values. We'll discuss function arguments in greater detail below.

\subsubsection{Creating vectors}

An important function that you've used extensively but we've glossed
over is the |c()| function. This is short for `concatenate' or
`combine' and as you've seen it combines it's arguments to form a
vector.

For vectors of more than 10 or so elements it gets tiresome and error
prone to create vectors using |c()|. For medium length vectors
the |scan()| function is very useful.
%
\begin{R}
> test.scores <- scan()
1: 98 92 78 65 52 59 75 77 84 31 83 72 59 69 71 66 
17: 
Read 16 items
> test.scores
 [1] 98 92 78 65 52 59 75 77 84 31 83 72 59 69 71 66
\end{R}
%
When you invoke |scan()| without any arguments the function
will read in a list of values separated by white space (usually spaces
or tabs). Values are read until |scan()| encounters a blank
line or the end of file (EOF) signal (platform dependent).

Note that we created a variable with the name |test.scores|.
If you have previous programming experience you might be surprised that
this works. Unlike most languages, R allows you to use periods in
variable names. Descriptive variable names generally improve readability
but they can also become cumbersome (e.g.
|my.long.and.obnoxious.variable.name|). As a general rule of
thumb use short variable names when working at the interpreter and more
descriptive variable names in functions.

\subsubsection{Useful Numerical Functions}

Let's introduce some additional numerical functions that are useful for
operating on vectors.
%
\begin{R}
> sum(test.scores)
[1] 1131
> min(test.scores)
[1] 31
> max(test.scores)
[1] 98
> range(test.scores) # min,max returned as a vec of len 2
[1] 31 98
> sorted.scores <- sort(test.scores)
> sorted.scores
 [1] 31 52 59 59 65 66 69 71 72 75 77 78 83 84 92 98
> w <- c(-1, 2, -3, 3)
> abs(w) # absolute value function
\end{R}


\subsection{Function Arguments in R}

Function arguments can specify the data that a function operates on or
parameters that the function uses. Some arguments are required, while
others are optional and are assigned default values if not specified.

Take for example the |log()| function. If you examine the help
file for the |log()| function you'll see that it takes two
arguments, refered to as `|x|' and `|base|'. The
argument |x| represents the numeric vector you pass to the
function and is a required argument (see what happens when you type
|log()| without giving an argument). The argument
|base| is optional. By default the value of |base|
is $e = 2.71828\ldots$. Therefore by default the |log()|
function returns natural logarithms. If you want logarithms to a
different base you can change the |base| argument as in the
following examples:
%
\begin{R}
> log(2) # log of 2, base e
[1] 0.6931472
> log(2,2) # log of 2, base 2
[1] 1
> log(2, 4) # log of 2, base 4
[1] 0.5
\end{R}



\subsection{Simple Input in R}

The \lstinline!c()! and \lstinline!scan()! functions are fine for
creating small to medium vectors at the interpreter, but eventually
you'll want to start manipulating larger collections of data. There are
a variety of functions in R for retrieving data from files.

The most convenient file format to work with are tab delimited text
files. Text files have the advantage that they are human readable and
are easily shared across different platforms. If you get in the habit of
archiving data as text files you'll never find yourself in a situation
where you're unable to retrieve important data because the binary data
format has changed between versions of a program.

\subsection{Using \lstinline!scan()! to input data}

\lstinline!scan()! itself can be used to read data out of a file.
Download the file \lstinline!algae.txt! from the class website and try
the following (after changing your working directory):

\begin{R}
> algae <- scan('algae.txt')
Read 12 items
> algae
 [1] 0.530 0.183 0.603 0.994 0.708 0.006 0.867 0.059 0.349 0.699 0.983 0.100
\end{R}
One of the things to be aware of when using \lstinline!scan()! is that
if the data type contained in the file can not be coerced to doubles
than you must specify the data type using the \lstinline!what! argument.
The \lstinline!what! argument is also used to enable the use of
\lstinline!scan()! with columnar data. Download \lstinline!algae2.txt!
and try the following:

\begin{R}
> algae.table <- scan('algae2.txt', what=list('',double(0))) 
                        # note use of list argument to what
> algae.table
[[1]]
 [1] "Jan" "Feb" "Mar" "Apr" "May" "Jun" "Jul" "Aug" "Sep" "Oct" "Nov" "Dec"

[[2]]
 [1] 0.530 0.183 0.603 0.994 0.708 0.006 0.867 0.059 0.349 0.699 0.983 0.100

> algae.table[[1]]
 [1] "Jan" "Feb" "Mar" "Apr" "May" "Jun" "Jul" "Aug" "Sep" "Oct" "Nov" "Dec"
> algae.table[[2]]
 [1] 0.530 0.183 0.603 0.994 0.708 0.006 0.867 0.059 0.349 0.699 0.983 0.100
\end{R}
Use help to learn more about \lstinline!scan()!.

\subsection{Using \lstinline!read.table()! to input data}

\lstinline!read.table()! (and it's derivates - see the help file)
provides a more convenient interface for reading tabular data. Using the
file \lstinline!turtles.txt!:

\begin{R}
> turtles <- read.table('turtles.txt', header=T)
> turtles
   sex length width height
1    f     98    81     38
2    f    103    84     38
3    f    103    86     42
  # output truncated
> names(turtles)
[1] "sex"    "length" "width"  "height"
> length(turtles)
[1] 4
> length(turtles$sex)
[1] 48  
\end{R}
What kind of data structure is \lstinline!turtles!? What happens when
you call the \lstinline!read.table()! function without specifying the
argument \lstinline!header=T!?

You'll be using the \lstinline!read.table()}!function frequently. Spend
some time reading the documentation and playing around with different
argument values (for example, try and figure out how to specify
different column names on input).

Note: \lstinline!read.table()! is more convenient but \lstinline!scan()!
is more efficient for large files. See the R documentation for more
info.

\subsection{Basic Statistical Functions in R}

There are a wealth of statistical functions built into R. Let's start to
put these to use.

If you wanted to know the mean carapace width of turtles in your sample
you could calculate this simply as follows:

\begin{R}
> sum(turtles$width)/length(turtles$width)
[1] 95.4375
\end{R}
Of course R has a built in \lstinline!mean()! function.

\begin{R}
mean(turtles$width) [1] 95.4375
\end{R}
One of the advantages of the built in \lstinline!mean()! function is
that it knows how to operate on lists as well as vectors:

\begin{R}
> mean(turtles)
      sex    length     width    height 
       NA 124.68750  95.43750  46.33333 
Warning message:
argument is not numeric or logical: returning NA in: mean.default(X[[1]], ...) 
\end{R}
Can you figure out why the above produced a warning message? Let's take
a look at some more standard statistical functions:

\begin{R}
> min(turtles$width)
[1] 74
> max(turtles$width)
[1] 132
> range(turtles$width)
[1]  74 132
> median(turtles$width)
[1] 93
> summary(turtles$width)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  74.00   86.00   93.00   95.44  102.00  132.00 
> var(turtles$width) # variance
[1] 160.6769
> sd(turtles$width)  # standard deviation
[1] 12.67584
\end{R}


\subsection{Simple Plots in R}

One of the advantages of R is it's ability to produce a variety of plots
and statistical graphics. Try out the following:

\begin{R}
> hist(turtles$width)  # histogram plot
> hist(turtles$width,10) # produces a histogram with 10 bins
> hist(turtles$width,breaks=10, xlab="Carapace Width", probability=T)
> boxplot(turtles$width) # simple box plot
# a fancy box plot showing multiple variables
> boxplot(list(turtles$length, turtles$width,  turtles$height),
+        names=c("Carapace\nLength","Carapace\nWidth","Carapace\nHeight"),
+        ylab="millimeters") 
> title("Turtle Shell Variables")
> plot(turtles$length, turtles$width)
# how does this differ from the plot above?
> plot(turtles$length ~ turtles$width) 
> plot(turtles$length, turtles$width, 
+      xlab="Carapace Length(mm)", ylab="Carapace Width(mm)")
> title("Relationship Between\nLength and Width")
\end{R}


To get a sense of some of the graphical power of R try the
\lstinline!demo()! function:
%
\begin{R}
> demo(graphics)
\end{R}
