
\documentclass[11pt,letterpaper]{article}

\usepackage[T1]{fontenc}
\usepackage[sc]{mathpazo}

\usepackage{indentfirst}   %% if you want to have the first paragraph of each section indented %%
\usepackage[top=0.75in, bottom=0.75in, left=1in, right=1in]{geometry} %% for setting page margins %%
\usepackage{hyperref}
\hypersetup{
colorlinks=true
}
\usepackage{titlesec} % for more compact section spacing
\titlespacing{\section}{0pt}{*2}{*1}

\begin{document}


\section*{\centering Biology 723: Scientific Computing for Biologists, Fall 2012}
\begin{center}
\begin{tabular}{ll}
\textbf{Instructor}: Paul M. Magwene & \textbf{Office}: FFSC 4103\\
\textbf{Phone}: 613-8159 & \textbf{Email}: paul.magwene@duke.edu
\end{tabular}
\end{center}

\section*{Description}

The focus of this course is statistical computing for the biological sciences with an emphasis on common multivariate statistical methods and techniques for exploratory data analysis. A major goal of the course is to help graduate students in the biological sciences develop practical insights into methods that they are likely to encounter in their own research, and the potential advantages and pitfalls that come with their use.

\section*{Prerequisites}

Enrollment is limited to graduate students or undergraduates with permission of instructor. No previous programming experience is required, but familiarity with basic statistical concepts (equivalent of STA 213) is assumed.

\section*{Grading}
Grading is based on weekly homework assignments. These homework assignments will typically consist of statistical problem solving exercises and/or programming tasks.

\section*{Course Website}

\href{https://github.com/pmagwene/Bio313}{https://github.com/pmagwene/Bio313}

\renewcommand{\refname}{Texts}
\begin{thebibliography}{99}

\bibitem{Wickens} Wickens, T.\ D. 1995. The geometry of multivariate statistics. Lawrence Earlbaum Associates, New Jersey.

\bibitem{Matloff} Matloff, N. 2011. The Art of R Programming. No Starch Press, San Francisco.

\bibitem{CSpy}A. B. Downey, J. Elkner  and C.\ Meyers. How to think like a computer scientist: learning with Python. Available in \href{http://www.greenteapress.com/thinkpython/html/index.html}{HTML} and \href{http://www.greenteapress.com/thinkpython/thinkpython.pdf}{PDF} form under an open source license.



\end{thebibliography}

\renewcommand{\refname}{Other Recommended Texts}

\begin{thebibliography}{99}
\setcounter{enumiv}{5}
\bibitem{Hamilton} Hamilton, A.\ G. 1989. Linear algebra: An introduction with concurrent examples. Cambridge University Press, Cambridge.

A well organized and readable introduction to linear algebra. This assumes no previous familiarity with lineage algebra.  You'll get maximum benefit from this text if you work through the short exercises that accompany each chapter.

\bibitem{Krzanowski} Krzanowski, W.\ J. 2000. Principles of multivariate analysis. Oxford Univ. Press, New York.

I would have made this a required text but it's become unreasonably expensive, even for used copies. Nonetheless, if you plan on having just on book on multivariate statistics on your bookshelf this is the one I'd recommend.

\bibitem{Sokal} Sokal, R.\ R. and F.\ J. Rohlf. 1995. Biometry. W.\ H. Freeman, New York.

Another good text to have on your bookshelf. A readable and well organized basic statistics book with examples drawn from the biological literature.

\end{thebibliography}

\section*{Syllabus}

\renewcommand{\arraystretch}{1.4}
\begin{center}
\begin{tabular}{rp{5.5in}}
\multicolumn{1}{c}{{\sl Date}} & \multicolumn{1}{c}{{\sl Topic}} \\

August 28 & Introduction; Getting Acquainted with R \\
September 4 & Data as vectors; Vector operations: dot product, correlation, regression as projection \\
September 11 & Linear Algebra Review I; Descriptive statistics as matrix operations\\
September 18 & Linear Algebra Review II; Regression models \\
September 25 & Eigenvectors and Eigenvalues; Principal Components Analysis \\
October 2 & Singular Value Decomposition, Biplots, and Correspondence Analysis\\
October 9 & Discriminant analysis and Canonical Variate Analysis\\
October 16 & \multicolumn{1}{c}{{\sc Fall Break}} \\
October 23 & Analyses based on Similarity/Distance I; Hierarchical and K-means clustering\\
October 30 & Analyses based on Similarity/Distance II; Multidimensional scaling\\
November 6 & Randomization and Monte Carlo Methods; Jackknife, Bootstrap, and Permutation Tests\\
November 13 & Building Simulations in Python\\
November 20 & Building Bioinformatics Pipelines I; Pipes, redirection, subprocesses \\
November 27 & Building Bioinformatics Pipelines II; Putting the concepts to work \\

& \multicolumn{1}{c}{{\sc Graduate Classes End November 30}} \\
\end{tabular}
\end{center}
%$^{*}$ We will need to reschedule the Nov. 30 class session as I will be travelling.


\end{document}
